A spiral galaxy's mass distribution uncovered through lensing and dynamics

### 2016-06-27 LensingPaper_draft_v4.01.tex
Rewrote the summary section.

\Wilma{[TO DO: The referee says: "The discussion section is too non-committal and speculative in parts. For instance,
in section 5.1, the M/Ls inferred from a number of the different models are compared
with M/Ls from the literature, but the authors do not make any value judgement
between the different models to come to any firmer conclusions about what they
believe the most convincing description of the IMF to be. At the moment, this
section seems to just be trying to show that their various models can be made to
agree with previous models from the literature, but it would be much more
interesting if some attempt could be made to resolve all the different models (both
in this paper and in previous works) or at least present them in a more coherent
framework, to try to come up with a better picture of the nature of J1331's IMF."]}

@@

%--------------------------------------------
\subsection{Summary}
%--------------------------------------------
\Wilma{[TO DO: Go through it and make it shorter.]}
%--------------------------------------------
The goal of this work was to constrain the matter distribution of the massive spiral galaxy J1331. J1331 is of special interest because it firstly has a large counter-rotating core and might therefore be a merger remnant and it secondly acts as strong gravitational lens for a background source. The latter allowed us to constrain the galaxy's mass from gravitational lensing. As second independent approach we used dynamical Jeans modelling to recover and disentangle the stellar and DM distribution. We focussed on the bulge region of J1331 to complement previous studies of J1331 by \citet{SWELLSIII} and \citet{SWELLSV}. The independent mass constraints which we derived from lensing and dynamics agree very well with each other within $R_\text{eff}$.

In our lensing approach we fitted a scale-free galaxy model to the lensing image position. This constrained the Einstein radius to within 2\% ($R_\text{ein}=(0.91\pm0.02)'' \hat{=}(1.83\pm0.04)~\text{kpc}$), and the Einstein mass to within 4\% ($M_\text{ein} = (7.8\pm0.3) \cdot 10^{10} M_\odot$), consistent with results by \citet{SWELLSIII}. The recovered mass distribution agrees also very well with the shape of the observed lensing images.

To get a handle on J1331's light distribution we used a MGE fit to J1331's surface brightness in the F814W filter in HST imaging by \citet{SWELLSI}. Based on this we found J1331's effective radius to be $R_\text{eff} \simeq 2.6'' \hat{=} 5.2~\text{kpc}$ and its total I-band luminosity to be $L_\text{I,tot} \simeq 5.6 \cdot 10^{10} L_\odot$.

We used axisymmetric JAM modelling to calculate the second velocity moment of a tracer and mass distribution given in terms of MGEs. The JAM model of the lens mass model gave an independent prediction which was consistent with observed stellar kinematics from major axis long slit spectroscopy by \citet{SWELLSV} within $R_\text{eff}$. We also fitted mass models with and without NFW halo to the stellar kinematics within $R'=3.5~\text{kpc}$. From this we deduced that a mass-follows-light model with constant total mass-to-light ratio and velocity anisotropy is not a good model for J1331---not even in the inner regions. To describe the observed stellar kinematics we either require a strong contribution of a DM halo already in the bulge, or a strongly varying stellar mass-to-light ratio. We also showed that it is possible to construct a model which includes the counter-rotating core and fits the rotation curve at large radii.

While both our independent mass models are consistent with each other within $\sim R_\text{eff}$, they do not agree with previous findings by \citet{SWELLSV}, nor the data at large radii. We attribute this to the fact that \citet{SWELLSV} derived mass constraints from kinematics in the outer regions only, while we focussed on the inner regions in detail. We speculate that a merger might have modified the kinematic structure and/or mass distribution of J1331 such that crucial model assumptions are not valid anymore. To resolve the ambiguities in J1331's mass distribution 2-dimensional kinematic maps of J1331 from integral-field unit spectroscopy would be needed.
%--------------------------------------------

>>>>

@@

%--------------------------------------------
\subsection{Summary}
%--------------------------------------------

%--------------------------------------------
\NEW{We constrained the matter distribution of the massive spiral galaxy J1331, which has a large counter-rotating core, probably due to a merger in its past, and acts as a strong gravitational lens for a background source. We used two independent methods to model J1331: gravitational lensing and dynamical Jeans modelling.}\OLD{The goal of this work was to constrain the matter distribution of the massive spiral galaxy J1331. J1331 is of special interest because it firstly has a large counter-rotating core and might therefore be a merger remnant and it secondly acts as strong gravitational lens for a background source. The latter allowed us to constrain the galaxy's mass from gravitational lensing. As second independent approach we used dynamical Jeans modelling to recover and disentangle the stellar and DM distribution.} We focussed on the bulge region of J1331 to complement previous studies of J1331 by \citet{SWELLSIII} and \citet{SWELLSV}. The \OLD{independent} mass constraints \OLD{which we derived} from lensing and dynamics agree very well with each other within $R_\text{eff}$.

In our lensing approach we fitted a scale-free galaxy model to the lensing image position. This constrained the Einstein radius to within 2\% ($R_\text{ein}=(0.91\pm0.02)'' \hat{=}(1.83\pm0.04)~\text{kpc}$), and the Einstein mass to within 4\% ($M_\text{ein} = (7.8\pm0.3) \cdot 10^{10} M_\odot$), consistent with results by \citet{SWELLSIII}.\OLD{The recovered mass distribution agrees also very well with the shape of the observed lensing images.}

\OLD{To get a handle on J1331's light distribution we used a MGE fit to J1331's surface brightness in the F814W filter in HST imaging by \citet{SWELLSI}. Based on this we found J1331's effective radius to be $R_\text{eff} \simeq 2.6'' \hat{=} 5.2~\text{kpc}$ and its total I-band luminosity to be $L_\text{I,tot} \simeq 5.6 \cdot 10^{10} L_\odot$.}\NEW{A MGE fit to J1331's surface brightness in the F814W filter in HST imaging by \citet{SWELLSI} helped us determining the effective radius, $R_\text{eff} \simeq 2.6'' \hat{=} 5.2~\text{kpc}$, and total I-band luminosity of the galaxy, $L_\text{I,tot} \simeq 5.6 \cdot 10^{10} L_\odot$.}

\NEW{Axisymmetric JAM modelling allowed a comparison between model predictions for the second velocity moment given a tracer and mass distribution and the observed stellar kinematics from major axis long slit spectroscopy by \citet{SWELLSV}.} \OLD{We used axisymmetric JAM modelling to calculate the second velocity moment of a tracer and mass distribution given in terms of MGEs.} The \NEW{independent} JAM model of the lens mass model \OLD{gave an independent prediction which }was consistent with observed\OLD{ stellar} kinematics\OLD{ from major axis long slit spectroscopy by \citet{SWELLSV}} within $R_\text{eff}$. We also fitted mass models with and without NFW halo to the stellar kinematics within $R'=3.5~\text{kpc}$. From this we deduced that a mass-follows-light model \NEW{(even} with \OLD{constant total mass-to-light ratio and }velocity anisotropy\NEW{)} is not a good model for J1331\OLD{---not even in the}\NEW{'s} inner regions. \NEW{This ruled out the previous findings of \citet{SWELLSV} for J1331's bulge. We discussed that, to}\OLD{ To} describe the observed stellar kinematics\NEW{,} we \OLD{either}\NEW{most likely} require a \OLD{strong}\NEW{moderate} contribution of a DM halo already in the bulge, \NEW{moderate tangential velocity anisotropy, $\beta_z \simeq -0.4\pm0.1$, }\OLD{or} \NEW{and possibly even} a \OLD{strongly} varying stellar mass-to-light ratio \NEW{within the bulge, which could be the result of the previous merger event}. \NEW{We argue that we expect the total stellar mass-to-light ratio within the bulge to be slightly less bottom-heavy than a Salpeter IMF, $\Upsilon_\text{I,*}\simeq 4.2\pm0.2$.} We also showed that it is possible to construct a model which includes the counter-rotating core and fits the rotation curve at large radii.

While both our independent mass models are consistent with each other within $\sim R_\text{eff}$, they do not \OLD{agree with previous findings by \citet{SWELLSV}, nor the}\NEW{describe the} data at large radii \NEW{very well}. \OLD{We attribute this to the fact that \citet{SWELLSV} derived mass constraints from kinematics in the outer regions only, while we focussed on the inner regions in detail.} We speculate \OLD{that}\NEW{how} a merger \OLD{might}\NEW{could} have modified the kinematic structure and/or mass distribution of J1331\OLD{ such that crucial model assumptions are not valid anymore}. To resolve the ambiguities in J1331's mass distribution 2-dimensional kinematic maps of J1331 from integral-field unit spectroscopy would be needed.
%--------------------------------------------
